### 2021-04-20 submission/manuscript.Rmd
Add to results

``` {r 5-day PEG microbiota}
f_d_permanova <- read_tsv(here("data/process/5_days_PEG_permanova_stools.tsv"))
f_d_sign_contrib <- f_d_permanova %>% arrange(desc(R2)) %>%
  filter(!effects == "Total") %>% #Remove total since it is not one of the variables
  filter(`Pr(>F)` < 0.05) #Only select significant contributors

@@

```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C. difficile* colonization dynamics between mice treated with laxatives versus clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy, we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B). Next, we examined  alpha diversity by looking at Shannon diversity index over time, although both clindamycin and PEG treatments decreased diversity, the diversity index was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig. 2C). We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-treatment. We found `r PEG_genera` OTUs that were altered by PEG, the majority of these OTUs decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment as Bacteroides actually decreased in clindamycin treated mice (Fig. 2D). Finally, we examined the bacteria that differ across treatment groups over multiple timepoints. We found... (Fig. 2E). Thus, PEG has a significant impact on the fecal microbiota that was maintained over time.

Surprisingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine what other bacteria were changing during the post-infection period when *C. difficile* changed from undectable to detectable. Interestingly, we found ___ bacteria were shifting during the time period when C. difficile was becoming detectable in the stools (Fig. S1)

**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given that previous work had demonstrated that PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity. We examined CDI severity by evaluating cecum and colon histopathology [].

***C. difficile* challenge does not have a synergistic disruptive effect on the microbiota of PEG-treated mice**
Because *C. difficile* itself can have an impact on the microbiota (ref), we also sequenced the tissue and stools of mock-challenged mice. Examining the stools of the mock-infected PEG- and clindamycin-treated mice revealed similar bacterial disruptions as the *C. difficile* challenged mice (Fig. S2A). Similarly, there was no difference between the tissues of mock and *C. difficile* challenged mice (Fig. S2B). Thus, most of the microbiota alterations we observed in the PEG-treated mice were a result of the laxative and not an interaction between the laxative and *C. difficile*.

>>>>

``` {r 5-day PEG fecal microbiota}
f_d_permanova <- read_tsv(here("data/process/5_days_PEG_permanova_stools.tsv"))
f_d_sign_contrib <- f_d_permanova %>% arrange(desc(R2)) %>%
  filter(!effects == "Total") %>% #Remove total since it is not one of the variables
  filter(`Pr(>F)` < 0.05) #Only select significant contributors

@@

#Genera that were different between treatment groups
f_d_group_genera <- read_tsv(here("data/process/5_days_PEG_genus_group_stools.tsv")) %>% 
  filter(p.value.adj < 0.05)  #Select only significant p-values
f_d_group_total_genera <- f_d_group_genera %>% 
  filter(!duplicated(genus))  %>% #Remove genera that appear multiple times
  tally()
#Genera that were different between treatment groups over multiple days
f_d_group_multi_day <- f_d_group_genera %>% 
  filter(p.value.adj < 0.05) %>% #Select only significant p-values
  group_by(genus) %>% 
  tally() %>% 
  filter(n > 1) %>% #select genera that vary over multiple timepoints = n > 1
  arrange(desc(n)) %>%  #Rank by how many time each genera shows up 
  nrow() #24 genera includes Peptostreptococcaceae and Unclassified

```
**5-day laxative treatment differentially disrupts the fecal microbiota compared to clindamycin treatment.** Since laxatives and clindamycin have previously been shown to disrupt the murine microbiota (ref), we hypothesized the different *C. difficile* colonization dynamics between mice treated with laxatives versus clindamycin were due to the two drugs having differential effects on the microbiota. We profiled the stool microbiota over time by sequencing the V4 region of the 16S rRNA gene to compare changes across treatment groups. We found time and treatment group explained half of the observed variation between fecal communities with most of the remaining variation explained by interactions between treatment group and other experimental variables including time, cage effects, and sequencing preparation (PERMANOVA combined R^2^ = `r f_d_top_R2_sum`, *P* `r format.pval(f_d_top_p, eps = 0.001)`, Fig. 2A, Data Set S1, Sheet X). Cage effects refer to the well-documented phenomenon that mice housed in the same cages have similar microbial communities due to copraphagy, we tried to minimize the impact of cage effects on our experiment by breaking up cagemates when assigning mice to treatment groups and primarily housing only two mice per cage. Importantly, although we conducted a total of `r five_day_total_exp` separate experiments, the experiment number and it's interaction with treatment group was not one of the variables that significantly explained the observed variation in fecal communities (Data Set S1, Sheet X). Interestingly, none of the treatment groups recovered to their baseline community structure either 10 or 30 days post-infection suggesting other community features besides recovery to baseline were responsible for the prolonged *C. difficile* colonization in PEG-treated mice (Fig. 2B).

Next, we examined  alpha diversity by looking at Shannon diversity index over time, although both clindamycin and PEG treatments decreased diversity, the diversity index was lower in the groups of mice that received PEG treatment compared to those that received clindamycin through thirty days post-infection (Fig. 2C). We next examined the bacterial genera that shifted after PEG treatment by comparing the baseline samples of mice treated with only PEG to samples from the same mice one day post-treatment. We found `r PEG_genera` OTUs that were altered by PEG, the majority of these OTUs decreased after PEG, but *Enterobacteriaceae* and *Bacteroides* increased and the increase in *Bacteroides* was unique to PEG treatment as Bacteroides actually decreased in clindamycin treated mice (Fig. 2D, Data Set S1, Sheet X). Finally, we examined the bacteria that differ across treatment groups over multiple timepoints. We found `r f_d_group_multi_day` were different over multiple timepoints out of the `r f_d_group_total_genera` that were different between treatment groups (Fig. 2E, Data Set S1, Sheet X). Thus, PEG has a significant impact on the fecal microbiota that was maintained over time and distinct from clindamycin treatment.

Surprisingly, *C. difficile* was not immediately detectable in the stools of the PEG-treated mice that were allowed to recover for 10 days prior to  challenge. We decided to examine the bacteria that changed during the post-infection period when the group median *C. difficile* shifted from undetectable at 1 day post-infection to detectable in the stool samples with the median stabilizing around 8 dayspost-infection (Fig. S1A). Interestingly, we found *Erysipelotrichaceae*, *Enterobacteriaceae*, and *Akkermansia* were changing during the time period when *C. difficile* was becoming detectable in the stools (Fig. S1B), although none of the bacteria were significant after multiple hypothesis correction (Data Set S1, Sheet X). Although we did not identify a clear signal to explain the delayed appearance of *C. difficile* in the 5-day PEG mice that were allowed to recover for 10 days prior to challenge, the delay is striking and could reflect changes in microbial activity or metabolites that were not examined in this study.

``` {r 5-day PEG tissue microbiota}
f_d_permanova_t <- read_tsv(here("data/process/5_days_PEG_permanova_tissues.tsv"))
f_d_sign_contrib_t <- f_d_permanova_t %>% arrange(desc(R2)) %>%
  filter(!effects == "Total") %>% #Remove total since it is not one of the variables
  filter(`Pr(>F)` < 0.05) #Only select significant contributors

#List of variables that significantly explained variation in stool samples
f_d_top_effects_t <- f_d_sign_contrib_t %>% 
  pull(effects)
#Sum of R2 values for variables that significantly explained variation in stool samples
f_d_top_R2_sum_t <- f_d_sign_contrib_t %>% 
  select(R2) %>% sum
#P-value for variables that significantly explained variation in stool samples
f_d_top_p_t <- f_d_sign_contrib_t %>% 
  select(`Pr(>F)`) %>% max # P value will be rounded in text

```
**5-day laxative treatment does not promote more severe CDIs despite altering the mucosal microbiota.** Given the findings from a previous study that demonstrated that PEG treatment disrupts the mucus layer and alters the immune response in mice [@Tropini2018], we decided to examine the impact of PEG treatment on the mucosal microbiota and CDI severity. To evaluate the mucosal microbiota, we sequenced snips of tissue collected from the cecum, proximal colon, and distal colon. Similar to what was observed with the stool samples, alpha diversity was lower in the PEG-treated mice compared to clindamycin treatment (Fig. 3A). Although alpha diversity continued to increase over time based on the communities from PEG-treated mice collected at 20 and 30 days post-infection (Fig. 3A). Group, day, and the interactions with other variables (cage, experiment number, and sample type) explained the majority of the variation observed in mucosal communities (PERMANOVA combined R^2^ = `r f_d_top_R2_sum_t`, *P* `r format.pval(f_d_top_p_t, eps = 0.05)`, Fig. 3B, Data Set S1, Sheet X). *Bacteroides*, *Butyricoccus*, *Clostridiales*, *Clostridium Cluster XIVb*, *Firmicutes*, and *Ruminococcaceae* were consistently different between treatment groups in both the tissue and stool communities (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Next, we examined CDI severity by evaluating cecum and colon histopathology [@Reeves2011] and found there was no difference in cecum and colon scores between clindamycin and PEG-treated mice that were challenged with *C. difficile* at 4 days post-infection (Fig. 3D), the timepoint typically examined in *C. difficile* 630 challenged mice (ref). We also looked at 6 days post-infection because that was when we started to see a large difference in C*. difficile* colonization levels between PEG- and clindamycin-treated mice (Fig. 1C). Although, there was a slight difference in the colon between PEG and clindamycin-treated mice, there was no difference in the cecum and the overall score is still relatively low given that the max possible summary score is 12 (Fig. 3E). Thus, although PEG treatment had a profound impact on the mucosal microbiota, the impact of *C. difficile* on the cecum and colon was similar between PEG and clindamycin treated mice.

***C. difficile* challenge does not have a synergistic disruptive effect on the microbiota of PEG-treated mice** Because *C. difficile* itself can have an impact on the microbiota [@Jenior2018], we also sequenced the tissue and stools of mock-challenged clindamycin and 5-day PEG treated mice. Examining the stools of the mock-infected PEG- and clindamycin-treated mice revealed similar bacterial disruptions as the *C. difficile* challenged mice (Fig. S2A-C). Similarly, there was no difference between the tissues of mock and *C. difficile* challenged mice (Fig. S2D-F). Thus, most of the microbiota alterations we observed in the PEG-treated mice were a result of the laxative and not an interaction between the laxative and *C. difficile*.

### 2021-04-20 submission/manuscript.tex
Add to results

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since laxatives and
clindamycin have previously been shown to disrupt the murine microbiota
(ref), we hypothesized the different \emph{C. difficile} colonization
dynamics between mice treated with laxatives versus clindamycin were due
to the two drugs having differential effects on the microbiota. We
profiled the stool microbiota over time by sequencing the V4 region of
the 16S rRNA gene to compare changes across treatment groups. We found
time and treatment group explained half of the observed variation
between fecal communities with most of the remaining variation explained
by interactions between treatment group and other experimental variables
including time, cage effects, and sequencing preparation (PERMANOVA
combined R\textsuperscript{2} = 0.95, \emph{P} \textless{} 0.001, Fig.
2A, Data Set S1, Sheet X). Cage effects refer to the well-documented
phenomenon that mice housed in the same cages have similar microbial
communities due to copraphagy, we tried to minimize the impact of cage
effects on our experiment by breaking up cagemates when assigning mice
to treatment groups and primarily housing only two mice per cage.
Importantly, although we conducted a total of 5 separate experiments,
the experiment number and it's interaction with treatment group was not
one of the variables that significantly explained the observed variation
in fecal communities (Data Set S1, Sheet X). Interestingly, none of the
treatment groups recovered to their baseline community structure either
10 or 30 days post-infection suggesting other community features besides
recovery to baseline were responsible for the prolonged \emph{C.
difficile} colonization in PEG-treated mice (Fig. 2B). Next, we examined
alpha diversity by looking at Shannon diversity index over time,
although both clindamycin and PEG treatments decreased diversity, the
diversity index was lower in the groups of mice that received PEG
treatment compared to those that received clindamycin through thirty
days post-infection (Fig. 2C). We next examined the bacterial genera
that shifted after PEG treatment by comparing the baseline samples of
mice treated with only PEG to samples from the same mice one day
post-treatment. We found 18 OTUs that were altered by PEG, the majority
of these OTUs decreased after PEG, but \emph{Enterobacteriaceae} and
\emph{Bacteroides} increased and the increase in \emph{Bacteroides} was
unique to PEG treatment as Bacteroides actually decreased in clindamycin
treated mice (Fig. 2D). Finally, we examined the bacteria that differ
across treatment groups over multiple timepoints. We found\ldots{} (Fig.
2E). Thus, PEG has a significant impact on the fecal microbiota that was
maintained over time.

Surprisingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine what other bacteria were
changing during the post-infection period when \emph{C. difficile}
changed from undectable to detectable. Interestingly, we found \_\_\_
bacteria were shifting during the time period when C. difficile was
becoming detectable in the stools (Fig. S1)

\textbf{5-day laxative treatment does not promote more severe CDIs
despite altering the mucosal microbiota.} Given that previous work had
demonstrated that PEG treatment disrupts the mucus layer and alters the
immune response in mice (16), we decided to examine the impact of PEG
treatment on the mucosal microbiota and CDI severity. We examined CDI
severity by evaluating cecum and colon histopathology {[}{]}.

\textbf{\emph{C. difficile} challenge does not have a synergistic
disruptive effect on the microbiota of PEG-treated mice} Because
\emph{C. difficile} itself can have an impact on the microbiota (ref),
we also sequenced the tissue and stools of mock-challenged mice.
Examining the stools of the mock-infected PEG- and clindamycin-treated
mice revealed similar bacterial disruptions as the \emph{C. difficile}
challenged mice (Fig. S2A). Similarly, there was no difference between
the tissues of mock and \emph{C. difficile} challenged mice (Fig. S2B).
Thus, most of the microbiota alterations we observed in the PEG-treated
mice were a result of the laxative and not an interaction between the
laxative and \emph{C. difficile}.

@@

>>>>

\textbf{5-day laxative treatment differentially disrupts the fecal
microbiota compared to clindamycin treatment.} Since laxatives and
clindamycin have previously been shown to disrupt the murine microbiota
(ref), we hypothesized the different \emph{C. difficile} colonization
dynamics between mice treated with laxatives versus clindamycin were due
to the two drugs having differential effects on the microbiota. We
profiled the stool microbiota over time by sequencing the V4 region of
the 16S rRNA gene to compare changes across treatment groups. We found
time and treatment group explained half of the observed variation
between fecal communities with most of the remaining variation explained
by interactions between treatment group and other experimental variables
including time, cage effects, and sequencing preparation (PERMANOVA
combined R\textsuperscript{2} = 0.95, \emph{P} \textless{} 0.001, Fig.
2A, Data Set S1, Sheet X). Cage effects refer to the well-documented
phenomenon that mice housed in the same cages have similar microbial
communities due to copraphagy, we tried to minimize the impact of cage
effects on our experiment by breaking up cagemates when assigning mice
to treatment groups and primarily housing only two mice per cage.
Importantly, although we conducted a total of 5 separate experiments,
the experiment number and it's interaction with treatment group was not
one of the variables that significantly explained the observed variation
in fecal communities (Data Set S1, Sheet X). Interestingly, none of the
treatment groups recovered to their baseline community structure either
10 or 30 days post-infection suggesting other community features besides
recovery to baseline were responsible for the prolonged \emph{C.
difficile} colonization in PEG-treated mice (Fig. 2B).

Next, we examined alpha diversity by looking at Shannon diversity index
over time, although both clindamycin and PEG treatments decreased
diversity, the diversity index was lower in the groups of mice that
received PEG treatment compared to those that received clindamycin
through thirty days post-infection (Fig. 2C). We next examined the
bacterial genera that shifted after PEG treatment by comparing the
baseline samples of mice treated with only PEG to samples from the same
mice one day post-treatment. We found 18 OTUs that were altered by PEG,
the majority of these OTUs decreased after PEG, but
\emph{Enterobacteriaceae} and \emph{Bacteroides} increased and the
increase in \emph{Bacteroides} was unique to PEG treatment as
Bacteroides actually decreased in clindamycin treated mice (Fig. 2D,
Data Set S1, Sheet X). Finally, we examined the bacteria that differ
across treatment groups over multiple timepoints. We found 24 were
different over multiple timepoints out of the 33 that were different
between treatment groups (Fig. 2E, Data Set S1, Sheet X). Thus, PEG has
a significant impact on the fecal microbiota that was maintained over
time and distinct from clindamycin treatment.

Surprisingly, \emph{C. difficile} was not immediately detectable in the
stools of the PEG-treated mice that were allowed to recover for 10 days
prior to challenge. We decided to examine the bacteria that changed
during the post-infection period when the group median \emph{C.
difficile} shifted from undetectable at 1 day post-infection to
detectable in the stool samples with the median stabilizing around 8
dayspost-infection (Fig. S1A). Interestingly, we found
\emph{Erysipelotrichaceae}, \emph{Enterobacteriaceae}, and
\emph{Akkermansia} were changing during the time period when \emph{C.
difficile} was becoming detectable in the stools (Fig. S1B), although
none of the bacteria were significant after multiple hypothesis
correction (Data Set S1, Sheet X). Although we did not identify a clear
signal to explain the delayed appearance of \emph{C. difficile} in the
5-day PEG mice that were allowed to recover for 10 days prior to
challenge, the delay is striking and could reflect changes in microbial
activity or metabolites that were not examined in this study.

\textbf{5-day laxative treatment does not promote more severe CDIs
despite altering the mucosal microbiota.} Given the findings from a
previous study that demonstrated that PEG treatment disrupts the mucus
layer and alters the immune response in mice (16), we decided to examine
the impact of PEG treatment on the mucosal microbiota and CDI severity.
To evaluate the mucosal microbiota, we sequenced snips of tissue
collected from the cecum, proximal colon, and distal colon. Similar to
what was observed with the stool samples, alpha diversity was lower in
the PEG-treated mice compared to clindamycin treatment (Fig. 3A).
Although alpha diversity continued to increase over time based on the
communities from PEG-treated mice collected at 20 and 30 days
post-infection (Fig. 3A). Group, day, and the interactions with other
variables (cage, experiment number, and sample type) explained the
majority of the variation observed in mucosal communities (PERMANOVA
combined R\textsuperscript{2} = 0.83, \emph{P} \textless{} 0.05, Fig.
3B, Data Set S1, Sheet X). \emph{Bacteroides}, \emph{Butyricoccus},
\emph{Clostridiales}, \emph{Clostridium Cluster XIVb},
\emph{Firmicutes}, and \emph{Ruminococcaceae} were consistently
different between treatment groups in both the tissue and stool
communities (Fig. 3C, Fig. 2E, Data Set S1, Sheet X). Next, we examined
CDI severity by evaluating cecum and colon histopathology (20) and found
there was no difference in cecum and colon scores between clindamycin
and PEG-treated mice that were challenged with \emph{C. difficile} at 4
days post-infection (Fig. 3D), the timepoint typically examined in
\emph{C. difficile} 630 challenged mice (ref). We also looked at 6 days
post-infection because that was when we started to see a large
difference in C\emph{. difficile} colonization levels between PEG- and
clindamycin-treated mice (Fig. 1C). Although, there was a slight
difference in the colon between PEG and clindamycin-treated mice, there
was no difference in the cecum and the overall score is still relatively
low given that the max possible summary score is 12 (Fig. 3E). Thus,
although PEG treatment had a profound impact on the mucosal microbiota,
the impact of \emph{C. difficile} on the cecum and colon was similar
between PEG and clindamycin treated mice.

\textbf{\emph{C. difficile} challenge does not have a synergistic
disruptive effect on the microbiota of PEG-treated mice} Because
\emph{C. difficile} itself can have an impact on the microbiota (21), we
also sequenced the tissue and stools of mock-challenged clindamycin and
5-day PEG treated mice. Examining the stools of the mock-infected PEG-
and clindamycin-treated mice revealed similar bacterial disruptions as
the \emph{C. difficile} challenged mice (Fig. S2A-C). Similarly, there
was no difference between the tissues of mock and \emph{C. difficile}
challenged mice (Fig. S2D-F). Thus, most of the microbiota alterations
we observed in the PEG-treated mice were a result of the laxative and
not an interaction between the laxative and \emph{C. difficile}.

@@

\leavevmode\hypertarget{ref-Reeves2011}{}%
20. \textbf{Reeves AE}, \textbf{Theriot CM}, \textbf{Bergin IL},
\textbf{Huffnagle GB}, \textbf{Schloss PD}, \textbf{Young VB}. 2011. The
interplay between microbiome dynamics and pathogen dynamics in a murine
model of \emph{Clostridium difficile} infection \textbf{2}:145--158.
doi:\href{https://doi.org/10.4161/gmic.2.3.16333}{10.4161/gmic.2.3.16333}.

\leavevmode\hypertarget{ref-Jenior2018}{}%
21. \textbf{Jenior ML}, \textbf{Leslie JL}, \textbf{Young VB},
\textbf{Schloss PD}. 2018. \emph{Clostridium difficile} alters the
structure and metabolism of distinct cecal microbiomes during initial
infection to promote sustained colonization. mSphere \textbf{3}.
doi:\href{https://doi.org/10.1128/msphere.00261-18}{10.1128/msphere.00261-18}.
